---
title: "MinneMUDAC Fall 2019"
author: "Muuzaani Nkhoma"
date: "Oct 12, 2019"
output: html_document
---

#### Load Packages run once before running the program or as needed

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(zoo)
library(car)
library(forecast)
library(fpp)

```




#####a. Read, Explore, and Prepare the data


```{r 1.a}

print("Start importing dataset")
treasurY <- read.csv("Data Time Series/Table B.1 US Treasury Security 10 Yr.csv", header = TRUE, skip = 3)
print("Finished importing dataset")

dim(treasurY)
ncol(treasurY)
nrow(treasurY)
names(treasurY)


```


##### View first observations

```{r }

head(treasurY)

```

##### View last observations
```{r }

tail(treasurY)

```

##### Summary Statistics

```{r }

summary(treasurY)

```

#####b. Create Times Series Data

```{r 1.b}

treasurY.ts <- ts(treasurY$Rate..., start = c(1953, 4), frequency = 12)

```

#####c. Plot Time Series Data

```{r 1.c}

plot.ts(treasurY.ts, type = "o", pch = 16, cex = .5, xlab = 'Year', 
        ylab = 'Rate', main = "Market Yield on US Treasury Securities at 10-year Constant Maturity")
#tsdisplay(cheese.ts)
```





#### b.Numerical Analysis
#####Plot ACF and PACF

```{r 2.b}

par(mfrow=c(1,2),oma=c(0,0,2,0))
acf(treasurY.ts,lag.max = 25,type = "correlation",  main = "ACF for Model #")
acf(treasurY.ts, lag.max = 25, type = "partial", main = "PACF for Model #")

```

#####c.1 Transformations
#####Check for volatility 

```{r 2c1}

treasurY.model<-lm(cheese$Production~cheese$Year) 
plot(cheese,type="l",xlab='Year',ylab='Production, 10000lb') 
points(cheese,pch=16,cex=.5) 
lines(cheese$Year, cheese.model$fit,col="red",lty=2) 
legend(1990,12000,c("Actual","Fits"), pch=c(16,NA),lwd=c(.5,.5),lty=c(1,2),cex=.55,col=c("black","red"))

```

####c.2.Residual Plots
#####Check for Nonconstant Variance(Heteroscedasticity)

```{r c.2}

par(mfrow=c(2,2),oma=c(0,0,0,0)) 
qqnorm(cheese.model$res,datax=TRUE,pch=16,xlab='Residual',main='Normal Q-Q Plot') 
qqline(cheese.model$res,datax=TRUE) 
plot(cheese.model$fit,cheese.model$res,pch=16, xlab='Fitted Value', ylab='Residual', main = 'Residuals vs Fitted Plot') 
abline(h=0) 
hist(cheese.model$res,col="gray",xlab='Residual',main='Residuals Histogram') 
plot(cheese.model$res,type="l",xlab='Observation Order', ylab='Residual', main = 'Residuals Plot') 
points(cheese.model$res,pch=16,cex=.5) 
abline(h=0)

```

####c.3.Box-Cox Transformation

```{r c.3}

lambda = BoxCox.lambda(cheese.ts)
cheese.model2 = BoxCox(cheese.ts, lambda=lambda)
plot.ts(cheese.model2, type = "o", pch = 16, cex = .5, xlab = 'Year', 
        ylab = 'Production', main = "Transformed Model of\n Us Production of Blue and Gorgonzola Cheese")
tsdisplay(cheese.model2)

```

####d.1 Trend and Seasonal Adjustments (Differencing)
#####Detect Trend and Seasonality
#####d.1.a Decomposition

#####d.1. Decomposition Plot

```{r d.1.b, eval=FALSE}

cheese.ts %>% decompose %>% autoplot

```

#####d.2 Detect Seasonality
#####d.2.a Seasonal Plot

```{r d.2.a, eval=FALSE}

seasonplot(cheese.ts)

```

#####d.2.b Monthly Plot

```{r d.2.b, eval=FALSE}

monthplot(cheese.ts)

```

#####d.2.c Differencing for Trend and Seasonality

```{r d.2.c, eval=FALSE}

monthplot(cheese.ts)

```

#####d.2.d Fit Seasonal ARIMA Model

```{r d.2.d, eval=FALSE}

monthplot(cheese.ts)

```

#####e. Trend and Seasonal Adjustments (Differencing)
#####Detect Trend and Nonstationarity

#####e.1 Non-Stationarity
#####Detect Non-Stationary Data

The stationarity of the data can be known by applying Unit Root Tests - Augmented Dickey-Fuller test (ADF), Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test.
Augmented Dickey-Fuller test (ADF)
The null-hypothesis for an ADF test is that the data are non-stationary. So p-value greater than 0.05 indicates non-stationarity, and  p-values less than 0.05 suggest stationarity.
KPSS Test
In this case, the null-hypothesis is that the data are stationary. In this case, p-value less than 0.05 indicates non-stationary series and p-value greater than 0.05 indicates stationary series.

```{r e.1}

# Unit Ratio Tests 
library(tseries)
adf = adf.test(cheese.ts) 
kpss = kpss.test(cheese.ts) 
adf
kpss

```

#####e.2 How to treat Non-Stationary Data

```{r e.2}

# Number of Differences Required to make data stationary 
ndiffs(cheese.ts)
cheese.ts1 = diff(cheese.ts, differences = 2)
plot.ts(cheese.ts1)

```




###SMOOTHING METHODS
####1.CONSTANT PROCESS-SIMPLE MOVING AVERAGE




#####d.Simple Moving Average Plot

```{r 1.d}

plot.ts(treasurY.ts, type = "o", pch = 16, cex = .5, xlab = 'Year', 
        ylab = 'Rate', main = "Market Yield on US Treasury Securities at 10-year Constant \nMaturity with Moving Average Plot")
treasurY.sma <- ma(treasurY.ts, order=12) # 12 month moving average
lines(treasurY.sma, col="red") # plot

```

####2.LEVEL PROCESS-SIMPLE EXPONENTIAL SMOOTHING

#####a. Read and prepare the data

```{r 2.a}

fatal <- read.csv("Data Time Series/Table B.25 U.S.Traffic Fatalities.csv")

```

#####b. Create and Plot Times Series Data

```{r 2.b}

fa.ts <- ts(fatal$Fatal, start = 1966, end = 2006)
plot(fa.ts, ylab = "Fatalities")

```


#####c. Simple Exponential Smoothing Plot

```{r 2.c}

fa.sm1 <- HoltWinters(fa.ts, beta = FALSE, gamma = FALSE)
plot(fa.sm1, ylab = "Fatalities")

```

#####d. one-year-ahead forecasts for 2007 - 2012

```{r 2.d}

fa.fore1 <- forecast(fa.sm1, h = 5)
plot(fa.fore1)
fa.fore1

```

#####e Model Performance
##### checks accuracy of forecast model

```{r 2.e}

fa.ac1 <- accuracy(fa.fore1)
fa.ac1

```

####3.LEVEL AND TREND PROCESS: DOUBLE EXPONENTIAL SMOOTHING

#####a. Double Exponential Smoothing Plot

```{r 3.a}

fa.sm2 <- HoltWinters(fa.ts, gamma = FALSE)
plot(fa.sm2, ylab = "Fatalities")

```

#####b. one-year-ahead forecasts for 2007 - 2012

```{r 3.b}

fa.fore2 <- forecast(fa.sm2, h = 5)
plot(fa.fore2)
fa.fore2

```

#####c. Model Performance
#####checks accuracy of forecast model

```{r 3.c}

fa.ac2 <- accuracy(fa.fore2)
fa.ac2

```

####4.LEVEL, TREND, and SEASONALITY PROCESS: HOLT-WINTERS


#####a. Read and prepare the data

```{r 4.a}

bevS <- read.csv("Data Time Series/Table B.5  US Beverage Manufacturer Product Shipments.csv", header = TRUE, skip = 3)

```


#####b. Create and Plot Times Series Data

```{r 4.b}

bevS.ts <- ts(bevS$Dollars..in.Millions, start = c(1992, 1), end = c(2006, 12), frequency = 12)
plot(bevS.ts, ylab = "Shipments")

```

#####c. Double Exponential Smoothing Plot-Additive

```{r 4.c}

bevS.sm1 <- HoltWinters(bevS.ts, seasonal = "additive")
plot(bevS.sm1, ylab = "Shipments")

```

#####d. Additive one-year-ahead forecasts for 2007 - 2012

```{r 4.d}

bevS.fore1 <- forecast(bevS.sm1, h = 5)
plot(bevS.fore1)
bevS.fore1

```

#####e. Model Performance-Additive
#####checks accuracy of forecast model

```{r 4.e}

bevS.ac1 <- accuracy(bevS.fore1)
bevS.ac1

```

#####f. Double Exponential Smoothing Plot-Multiplicative

```{r 4.f}

bevS.sm2 <- HoltWinters(bevS.ts, seasonal = "multiplicative")
plot(bevS.sm2, ylab = "Shipments")

```


####5.FIRST-ORDER EXPONENTIAL SMOOTHING

#####a. Read and prepare the data

```{r 5.a}

fatal <- read.csv("Data Time Series/Table B.25 U.S.Traffic Fatalities.csv")

```

#####b. Create and Plot Times Series Data

```{r 5.b}

fatal.ts <- ts(fatal$Fatal, start = 1966, end = 2006)
plot(fa.ts, ylab = "Fatalities")

```
#####c. First Order Exponential Smoothing Function

```{r 5.c1}

firstsmooth<-function(y,lambda,start=y[1])
              { 
                ytilde <- y 
                ytilde[1] <- lambda*y[1]+(1-lambda)*start 
                for (i in 2:length(y))
                  { 
                    ytilde[i]<-lambda*y[i]+(1-lambda)*ytilde[i-1] 
                  }
                ytilde 
              }

```

#####c. First Order Exponential Smoothing Plot

```{r 5.c}

fatal.sm1 <- firstsmooth(y = fatal.ts,lambda=0.4)
plot(fatal.ts,type="p", pch=16, cex=.5,   xlab='Date', ylab = "Fatalities",  xaxt='n')
lines(fatal.sm1)

```

#####d. one-year-ahead forecasts for 2007 - 2012

```{r 5.d}

fatal.fore1 <- forecast(fatal.sm1, h = 5)
plot(fatal.fore1)
fatal.fore1

```

#####e. Model Performance
#####checks accuracy of forecast model

```{r 5.e}

fatal.ac1 <- accuracy(fatal.fore1)
fatal.ac1

```

####6.SECOND-ORDER EXPONENTIAL SMOOTHING

#####a. Second-Order Exponential Smoothing Plot

```{r 6.a}

fatal.sm2 <- firstsmooth(y = fatal.sm1,lambda=0.4)
plot(fatal.ts,type="p", pch=16, cex=.5,   xlab='Date', ylab = "Fatalities",  xaxt='n')
lines(fatal.sm2)

```

#####d. one-year-ahead forecasts for 2007 - 2012

```{r 6.d}

fatal.fore2 <- forecast(fatal.sm2, h = 5)
plot(fatal.fore2)
fatal.fore2

```

#####e. Model Performance
#####checks accuracy of forecast model

```{r 6.e}

fatal.ac2 <- accuracy(fatal.fore2)
fatal.ac2

```

